Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2019-10-28 manuscript/abstract.tex
Address some open questions and comments about terminology
Addressed open questions where Trevor signed off on existing content and also
question about substitution-based metrics where we have added supplemental
analyses to address the question. This commit also updates references to a
"fitness model" instead of an "exponential growth model".

\begin{itemize}
\item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
\item{Should we calculate pairwise distances using a more evolutionarily-informed metric than just Hamming distance? \tbc{I'm quite happy with Hamming distance.}}
\item{Why do all of the substitution-based fitness metrics perform so poorly (especially the epitope mutations) when epitope mutations have been previously associated with viral fitness? \tbc{This is still concerning. I wonder about doing a quick calculation of a tree that we output to count epitope mutations and non-epitope mutations to trunk vs side-branches. We should be able to replicate the previous results. Katherine was able to do this just now with modern data with the Wolf epitope sites.}}
\item{How does our model performance compare with the original clade-based models? \tbc{I think the clade performance figure is enough here.}}
\item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}
\end{itemize}

>>>>

\begin{itemize}
\item{How should forecast uncertainty be calculated? \tbc{Important and will be commented on by reviewers. I think this can be carried by purely retrospective comparisons however.}}
\item{\tbc{Do we need to calculate LBI via a region-weighted scheme? May be better left for a later update? But seems important given centrality of LBI to our predictions.}}
\end{itemize}

### 2019-10-28 manuscript/main.tex
Address some open questions and comments about terminology
Addressed open questions where Trevor signed off on existing content and also
question about substitution-based metrics where we have added supplemental
analyses to address the question. This commit also updates references to a
"fitness model" instead of an "exponential growth model".

Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the exponential growth model (Fig.~\ref{fig:model}).
\tbc{I'd generally use the term `fitness model' instead of `exponential growth' model. Exponential growth makes me think of actual exponential growth rather than Malthusian growth.}
With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
\rnc{would be good to compare the year-on-year distance to diversity. that would give some sense whether a 2aa reduction is big or small.}
As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.16 $\pm$ 1.19 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection}).
With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
LBI was the best of these models, reducing the distance between populations by 1.41 $\pm$ 1.30 amino acids on average.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 $\pm$ 0.58 amino acids.
Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of samples under our simulated conditions was dominated by purifying selection rather than by positive selection at epitope sites.

@@

We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by an exponential growth model, as previously described \cite{Luksza:2014hj}.
Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.

>>>>

Here, we present a model of seasonal influenza evolution inspired by the Malthusian growth fitness model of {\L}uksza and L\"assig \cite{Luksza:2014hj}.
As with this original model, we seek to forecast the frequencies of viral populations one year in advance by applying to each virus sample an exponential growth factor scaled by an estimate of the sample's fitness (Fig.~\ref{fig:model}).
We estimate the frequency of virus samples every six months using a smoothed KDE kernel to represent the frequency of each sample.
We estimate viral fitness with biologically-informed metrics including those originally defined by \cite{Luksza:2014hj} of epitope cross-immunity and non-epitope mutations as well as four more recent metrics including hemagglutination inhibition (HI) cross-immunity \cite{Neher:2016hy}, deep mutational scanning (DMS) mutational effects \cite{Lee2018}, local branching index (LBI) \cite{Neher:2014eu}, and change in clade frequency over time (delta frequency).
We fit models by learning coefficients for each fitness metric either individually or in linear combinations from training data and select the best of these models using time-series cross-validation.
After selecting optimal models from training and validation, we evaluate the true out-of-sample errors of these models on additional data that were held out from the initial model fitting and tuning.
Importantly, our models find fitness coefficients that minimize the normalized average Hamming distance between the observed population one year in the future and the estimated population produced by the fitness model (Fig.~\ref{fig:model}).
With this approach, we avoid the intrinsic instability of clade definitions due to variability in phylogenetic reconstruction from year to year.
However, we retain the benefits of fitting models to highly similar samples found within clades and enable future forecasting efforts for pathogens whose sequences are not amenable to standard phylogenetic inference.

@@

The average distance between yearly populations, as measured by the naive model, was 8.97 $\pm$ 1.35 amino acids (Supplemental Fig.~\ref{sup_fig:distance_of_simulated_populations_between_timepoints}).
\rnc{would be good to compare the year-on-year distance to diversity. that would give some sense whether a 2aa reduction is big or small.}
As expected, the true fitness model outperformed all other models reducing the distance between populations by 2.16 $\pm$ 1.19 amino acids on average and surpassing the naive model in 32 of 33 (97\%) timepoints (Table~\ref{table_simulated_model_selection}).
With the exception of epitope cross-immunity, all biologically-informed models consistently outperformed the naive model (Fig.~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}).
LBI was the best of these models, reducing the distance between populations by 1.41 $\pm$ 1.30 amino acids on average.
Indeed, both growth-based models received positive coefficients and outperformed the mechanistic models.
The non-epitope mutations metric received a consistently negative coefficient with an average improvement of 0.71 $\pm$ 0.58 amino acids.
Surprisingly, the composite model of epitope cross-immunity and non-epitope mutations did not perform better than the individual non-epitope mutations model (Supplemental Fig.~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
While epitope cross-immunity was not predictive on its own, the consistently positive coefficient it received in the composite model suggested that it was an informative metric.
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated samples was dominated by purifying selection rather than by positive selection at epitope sites.

@@

We assumed that the evolution seasonal influenza A/H3N2 populations can be represented by a Malthusian growth fitness model, as previously described \cite{Luksza:2014hj}.
Under this model, we estimated the future frequency of the global population, $\mathbf{\hat{x}}$, at some time in the future, $t + \Delta{t}$, based on the current frequency, $x_{i}(t)$, and fitness, $f_{i}(t)$, of each sample $i$ as follows where the resulting future frequencies were normalized to one by $\frac{1}{Z(t)}$.
